---
title: "sobol_ode_SIBLEY"
format: html
---

## Set Up

```{r}
# Libraries 
library(here)
library(sensitivity)
library(purrr)
library(tidyverse)

# Source the forest growth function 
source(here("assignment_ode", "forest_growth.R"))
```






Consider the following model of forest growth (where forest size in measured in units of carbon (C))
- dC/dt = r * C for forests where C is below a threshold canopy closure
- dC/dt = g* 1( - C/K) for forests where carbon is at or above the threshold canopy closure
- K is a carrying capacity in units of carbon
The size of the forest (C), Canopy closure threshold and carrying capacity are all in units of carbon You could think of the canopy closure threshold as the size of the forest at which growth rates change from exponential to linear You can think of r, as early exponential growth rate and gas the linear growth rate once canopy closure has been reached

1. Implement this model in R (as a differential equation)
```{r}
library(tidyverse)
library(deSolve)
library(sensitivity)
library(here)

source(here("R/forest_growth.R"))
```

2. Run the model for 300 years (using the ODE solver) starting with an initial forest size of 10 kg/C, and using the following parameters:
  - canopy closure threshold of 50 kgC
  - K = 250 kg C (carrying capacity)
  - r = 0.01 (exponential growth rate before before canopy closure)
  - g = 2kg/year (linear growth rate after canopy closure)
  
```{r}
# Initial forest size (kg/C)
C0 <- c(C=10)

# Gets results for 300 years (evaluating every year)
simtimes <- seq(from = 1, to = 300)

# List of parameters
parms <- list(thresh = 50, K = 250, r = 0.01, g = 2)

# Run the model
result <- ode(y = C0, times = simtimes, func = forest_growth, parms = parms)
```

3. Graph the results. Here you are graphing the trajectory with the parameters as given (eg no uncertainty)

```{r}
ggplot(result, aes(time, C)) +
  geom_line() +
  geom_hline(yintercept = 50) + # canopy threshold
  labs(x = "Year",
       y = "Size of the forest (kg of carbon)",
       title = "Model of forest growth") +
  theme_minimal()
```

4. Run a sobol global (vary all parameters at the same time) sensitivity analysis that explores how the estimated maximum forest size (e.g maximum of C 300 years, varies with these parameters
- pre canopy closure growth rate (r)
- post-canopy closure growth rate (g)
- canopy closure threshold and carrying capacity (K)
Assume that parameters are al normally distributed with means as given above and standard deviation of 10% of mean value

```{r}
# Set initial value
C0 <- 10

# Create parameter distribution --> standard deviation is 10% of mean value
np <- 2000
thresh <- rnorm(mean = 50, sd = 5, n = np)
K <- rnorm(mean = 250, sd = 25, n = np)
r <- rnorm(mean = 0.01, sd = 0.001, n = np)
g <- rnorm(mean = 2, sd = 0.2, n = np)
X1 <- cbind.data.frame(r = r, K = K, g = g, thresh = thresh)

# Create second parameter distribution --> standard deviation is 10% of mean value
np <- 2000
thresh <- rnorm(mean = 50, sd = 5, n = np)
K <- rnorm(mean = 250, sd = 25, n = np)
r <- rnorm(mean = 0.01, sd = 0.001, n = np)
g <- rnorm(mean = 2, sd = 0.2, n = np)
X2 <- cbind.data.frame(r = r, K = K, g = g, thresh = thresh)

# fix any negative values and they are not meaningful
X1 <- X1 %>% map_df(pmax, 0.0)
X2 <- X2 %>% map_df(pmax, 0.0)

```

```{r}
# Run the sobol analysis
sens_C <- sobolSalt(model = NULL, X1, X2, nboot = 300)

# lets add names
colnames(sens_C$X) <- c("r", "K", "g", "thresh")
head(sens_C$X)

# Make the parameters into a list to put into function
params <- list(r = sens_C$X[1, "r"], g = sens_C$X[1, "g"], K = sens_C$X[1, "K"], thresh = sens_C$X[1, "thresh"])

result <- ode(y = C0, times = simtimes, func = forest_growth, parms = params)
result <- as.data.frame(result)

# Run sensitivity analysis
sens_C <- sensitivity::tell(sens_C, result, result.names = "ga")
sens_C
```

5. Graph the results of the sensitivity analysis as a box plot of maximum forest size and record the two Sobol indices (S and T).



6. In 2-3 sentences, discuss what the results of your simulation might mean. (For example think about how what parameters climate change might influence).

Submit Quarto with model implementation, graphs and sensitivity analysis and R file with your model